---
title: "Skewness-Based Feature QC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Skewness-Based Feature QC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Upper limit of quantification (skewness)

Some measurement platforms have upper limits of quantification which can result in skewed data with ceiling effects. In this vignette we:

1. simulate a dataset (`1000` samples x `500` features),
2. show 3 feature types (normal, left-skew, right-skew),
3. explain what skewness means,
4. run skewness filtering in `quality_control()`,
5. and visualize pre/post filtering impact.

```{r setup}
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(".")
} else {
  library(metaboprep)
}
library(ggplot2)

set.seed(20260408)
```


## 1) Simulate Data (1000 Samples x 500 Features)

We create 3 feature groups:

- `normal`: approximately symmetric distributions,
- `left_skew`: features with upper-end truncation (mimicking ceiling effects),
- `right_skew`: positively skewed features.

```{r simulate_data}
n_samples <- 1000
n_normal <- 380
n_left_skew <- 80
n_right_skew <- 40
stopifnot(n_normal + n_left_skew + n_right_skew == 500)

# 1) approximately symmetric features
normal_block <- matrix(
  rnorm(n_samples * n_normal, mean = 0, sd = 1),
  nrow = n_samples,
  ncol = n_normal
)

# 2) left-skew features generated via upper-end truncation
left_raw <- matrix(
  rnorm(n_samples * n_left_skew, mean = 1.2, sd = 0.9),
  nrow = n_samples,
  ncol = n_left_skew
)
upper_limit <- 0.7
left_skew_block <- pmin(left_raw, upper_limit)

# 3) right-skew features
right_block <- matrix(
  rlnorm(n_samples * n_right_skew, meanlog = 0, sdlog = 0.55),
  nrow = n_samples,
  ncol = n_right_skew
)

data <- cbind(normal_block, left_skew_block, right_block)

colnames(data) <- c(
  paste0("normal_", sprintf("%03d", seq_len(n_normal))),
  paste0("left_skew_", sprintf("%03d", seq_len(n_left_skew))),
  paste0("right_skew_", sprintf("%03d", seq_len(n_right_skew)))
)
rownames(data) <- paste0("sample_", seq_len(n_samples))

samples <- data.frame(sample_id = rownames(data))
features <- data.frame(
  feature_id = colnames(data),
  simulated_group = c(
    rep("normal", n_normal),
    rep("left_skew", n_left_skew),
    rep("right_skew", n_right_skew)
  )
)
```

Representative distributions for one feature from each group:

```{r example_histograms, fig.width=10, fig.height=4.8}
inspect_features <- c("normal_001", "left_skew_001", "right_skew_001")
plot_df <- data.frame(
  feature = rep(inspect_features, each = n_samples),
  value = as.vector(data[, inspect_features, drop = FALSE])
)

ggplot(plot_df, aes(x = value)) +
  geom_histogram(bins = 35, fill = "#377EB8", color = "white") +
  facet_wrap(~feature, scales = "free", nrow = 1) +
  theme_classic() +
  labs(
    x = "Feature abundance value",
    y = "Count (samples)",
    title = "Example feature distributions by simulated type"
  )
```

This is the key motivation: some individual **features** have visibly skewed distributions. In the above we can see how the distribution of feature `001` changes based on the simulation; we will focus on the `left_skew` data.


## 2) What skewness means (per feature)

`metaboprep` computes skewness **for each feature across samples** using `psych::describe()` and takes the `skew`
column (default `type = 3` in `psych::skew`):

\[
\mathrm{skew} = \frac{\sum_{i=1}^{n} (x_i-\bar{x})^3}{n \cdot s^3}
\]

where \(x_i\) are sample values for one feature, \(\bar{x}\) is the feature mean,
\(s\) is the feature SD, and \(n\) is the number of non-missing sample values.

Plain-language interpretation:

- `skew < 0` (`left_skew`): most values are concentrated on the right side, with a tail extending to the left.
- `skew > 0` (`right_skew`): most values are concentrated on the left side, with a tail extending to the right.
- larger `abs(skew)` means stronger asymmetry.

### What the threshold means

The threshold is a **cutoff on skewness magnitude**, not a p-value.

- With `direction = "left"`, a feature is excluded if `skew <= -threshold`.
- With `direction = "right"`, a feature is excluded if `skew >= threshold`.
- With `direction = "both"`, a feature is excluded if `abs(skew) >= threshold`.

So for `threshold = 1.25` and `direction = "left"`, features with skewness
`<= -1.25` are flagged/excluded.

The helper `feature_skewness()` returns one row per feature:

```{r feature_skewness_estimation}
skew_df <- feature_skewness(
  data = data,
  threshold = 1.25,
  direction = "left"
)
skew_df$simulated_group <- features$simulated_group[match(skew_df$feature_id, features$feature_id)]

head(skew_df[order(skew_df$skew), ], 10)
```

Skewness distributions by simulated group; we can see that the normally distributed features are mostly around `skew = 0`, the left-skewed features have negative skewness values (many below `-1.25`), and the right-skewed features have positive skewness values.

```{r skew_by_group_plot, fig.width=9.5, fig.height=4.8}
ggplot(skew_df, aes(x = simulated_group, y = skew, fill = simulated_group)) +
  geom_boxplot(outlier.alpha = 0.25) +
  theme_classic() +
  scale_fill_manual(values = c(
    normal = "#9ECAE1",
    left_skew = "#E41A1C",
    right_skew = "#4DAF4A"
  )) +
  guides(fill = "none") +
  labs(
    x = "Simulated feature type",
    y = "Skewness value (per feature)",
    title = "Feature-level skewness by simulated group"
  )
```


## 3) Apply a skewness filtering rule

To keep this vignette fast while using `1000 x 500` data, we apply the same
feature-level rule directly:

```{r apply_skew_rule}

skew_rule <- feature_skewness(
  data = data,
  threshold = 1.25,
  direction = "left"
)

excluded_skew <- skew_rule$feature_id[skew_rule$exclude_by_skewness %in% TRUE]
```

How many features are excluded by skewness?

```{r excluded_features}
length(excluded_skew)
head(excluded_skew, 20)
```

Exclusions by simulated feature type:

```{r excluded_summary}
excluded_tbl <- data.frame(feature_id = excluded_skew)
excluded_tbl$simulated_group <- features$simulated_group[match(excluded_tbl$feature_id, features$feature_id)]
table(excluded_tbl$simulated_group, useNA = "ifany")
```

Equivalent `quality_control()` call in `metaboprep`:

```{r qc_call_reference, eval=FALSE}
m <- Metaboprep(data = data, samples = samples, features = features)

m_qc <- quality_control(
  metaboprep = m,
  source_layer = "input",
  sample_missingness = 0.2,
  feature_missingness = 0.2,
  feature_skewness_threshold = 1.25,
  feature_skewness_direction = "left",
  total_peak_area_sd = NA,
  outlier_udist = 5,
  pc_outlier_sd = NA
)
```


## 4) Post-filtering impact on distributions

All plots below are feature-level summaries (counting features), except where explicitly stated.

```{r post_filter_prepare}
included_features <- setdiff(colnames(data), excluded_skew)

pre_skew <- feature_skewness(data, threshold = NULL)

post_matrix <- data[, included_features, drop = FALSE]
post_skew <- feature_skewness(post_matrix, threshold = NULL)

pre_skew$stage <- "Pre-filter"
post_skew$stage <- "Post-filter"
skew_compare <- rbind(
  pre_skew[, c("feature_id", "skew", "stage")],
  post_skew[, c("feature_id", "skew", "stage")]
)
```

### 4a) Feature-skewness distribution before vs after filtering

```{r post_filter_skew_hist, fig.width=10, fig.height=4.8}
ggplot(skew_compare, aes(x = skew, fill = stage)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.85) +
  facet_wrap(~stage, nrow = 1) +
  geom_vline(xintercept = 0, color = "#4D4D4D", linetype = "dashed") +
  geom_vline(xintercept = -skew_threshold, color = "#E41A1C", linetype = "dashed", linewidth = 1) +
  theme_classic() +
  scale_fill_manual(values = c("Pre-filter" = "#9ECAE1", "Post-filter" = "#3182BD")) +
  guides(fill = "none") +
  labs(
    x = "Skewness value (per feature)",
    y = "Count (features)",
    title = "Feature-level skewness distribution before and after filtering",
    subtitle = paste0("Red dashed line: left-skew exclusion threshold = ", -skew_threshold)
  )
```

### 4b) Feature retention by simulated type

```{r post_filter_group_counts, fig.width=9, fig.height=4.8}
pre_counts <- as.data.frame(table(features$simulated_group), stringsAsFactors = FALSE)
names(pre_counts) <- c("simulated_group", "n")
pre_counts$stage <- "Pre-filter"

post_counts <- as.data.frame(table(features$simulated_group[features$feature_id %in% included_features]), stringsAsFactors = FALSE)
names(post_counts) <- c("simulated_group", "n")
post_counts$stage <- "Post-filter"

counts_df <- rbind(pre_counts, post_counts)
counts_df$simulated_group <- factor(counts_df$simulated_group, levels = c("normal", "left_skew", "right_skew"))

ggplot(counts_df, aes(x = simulated_group, y = n, fill = stage)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  theme_classic() +
  scale_fill_manual(values = c("Pre-filter" = "#9ECAE1", "Post-filter" = "#3182BD")) +
  labs(
    x = "Simulated feature type",
    y = "Count (features)",
    title = "Feature retention by simulated type after skewness filtering"
  )
```

### 4c) Skewness profiles by feature type (pre vs post)

```{r post_filter_group_skew_boxplot, fig.width=10, fig.height=5}
pre_plot <- merge(pre_skew, features, by = "feature_id", all.x = TRUE)
post_plot <- merge(post_skew, features, by = "feature_id", all.x = TRUE)
box_df <- rbind(
  data.frame(stage = "Pre-filter", simulated_group = pre_plot$simulated_group, skew = pre_plot$skew),
  data.frame(stage = "Post-filter", simulated_group = post_plot$simulated_group, skew = post_plot$skew)
)

box_df$simulated_group <- factor(box_df$simulated_group, levels = c("normal", "left_skew", "right_skew"))

ggplot(box_df, aes(x = simulated_group, y = skew, fill = stage)) +
  geom_boxplot(outlier.alpha = 0.25, position = position_dodge(width = 0.75)) +
  geom_hline(yintercept = -skew_threshold, color = "#E41A1C", linetype = "dashed") +
  theme_classic() +
  scale_fill_manual(values = c("Pre-filter" = "#9ECAE1", "Post-filter" = "#3182BD")) +
  labs(
    x = "Simulated feature type",
    y = "Skewness value (per feature)",
    title = "Shift in feature-level skewness profiles after filtering"
  )
```

